Index measures for oak decline severity using phenotypic descriptors

### 2019-12-12 manuscript/manuscript.Rmd
added to methods

---
title: "Index measures for oak decline severity using phenotypic descriptors"
output: 
  #bookdown::word_document2:
  # toc: false
  bookdown::pdf_document2: default
bibliography: references.bib
---

@@

**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning

@@

- Site types
- Field phenotyping protocol
- Crown assessments
- Other assessments

@@

A total of 30 phenotypic descriptors were collected relating to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
These descriptors are shown in Table #.
Crown radius, canopy closure and tap tests for dead stem tissue were measured for each cardinal point.
The crown radius was
The AOD stem symptoms (active bleeds, black staining, callused wounds) were collected by measuring longest axis of each individual bleed, stain or callused wound found within the basal 3m of the stem (ref).
The frequency of *A. biguttatus* exit holes were counted along with the presence or absence of each of oval shaped exit holes, small circular shaped exit holes and ground level fungal fruiting bodies were recorded from the basal 2m of the stem (ref). 
All trees were manually assigned a decline status category; control, AOD, COD or remission.

@@

- Descriptors converted to presence and absence
- Aggregation of cardinal observations.
- Symptom type sizes average and frequency taken
- Random forest and MDS analyses
- Descriptions of collected phenotypic descriptors and manual status assignment

R v`r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and subsequent analysis of phenotypic data [@R2019].
The code used to generate the decline indexes is available as an R package pdi v`r packageVersion('pdi') %>% as.character()` at https://github.com/jasenfinch/pdi.
All raw data and code used to generate this manuscript are available at https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors.

@@

Four additional composite descriptors were calculated to account for the three dimensional attributes of the tree and covariate nature of certain descriptors as suggested by @@zarnoch2004.
These included:

@@

***Agrilus* exit hole density (m^-2^)** - density of observed *Agrilus* exit holes across the surveyed 2m basal trunk area.
$$e = \frac{h}{\frac{2d}{100}\pi}$$
$h$ = number of *Agrilus* exit holes, $d$ = diameter at breast height (cm)

@@

To account for these differences and to make the trees across the sites comparable in the context of decline, an adjustment of site means to the overall mean was applied to each of the size descriptors (diameter at breast height, lower crown height, timber height, total height, crown radius). 
An example of this adjustment is shown for diameter at breast height in Supplementary Figure # and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table #.  
This adjustment removed inter-site variability but maintained intra-site variability, allowing comparisons to be made between the trees at the different sites in the context of decline severity.

@@

The phenotypic descriptors were analysed using 100 repetitions of unsupervised random forest using the randomForest R package v`r packageVersion('randomForest') %>% as.character()` [@liaw2002].
The proximity 
The relative proximities and the phenotypic descriptors multidimensional scaling (MDS) to two dimensions was performed to yield two phenotypic indexes:

Proximity matrix (*P*) was transformed to a dissimilarity (1 - *P*) then multidimensional scaling (MDS) to 2 dimensions applied.
The first dimension was found to be related to the overall severity of decline.
It was min-max scaled to give the PDI a scale between 0 and 1 with more severely declining trees having a score closer to 1.
The second MDS dimension was found to be associated with  min-max scaled to give scores between -1 and 1 with chronically declining trees having a score closer to -1 and acutely declining trees having a score closer to 1. 
Trees with a score around 0 are neither distinctively AOD or COD sharing characteristics of both conditions in trees with PDI greater than 0.3.

@@

Individual random forest regression models of the phenotypic descriptors were trained, supervised using each decline index.
Optimal number of trees (ntree) and number randomly of variables (mtry) parameters for random forest were tuned prior to training. 
The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and as follows: PDI - `r glue(ntree = {PDI_rf_tune_params$ntree}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)})`; DAI - ntree = , mtry = , PMAE = (Supplementary figure #)
To understand the descriptors contributing to each of the indexes, Local Interpretable Model-Agnostic Explanations (LIME) using the lime R package v`r packageVersion('lime') %>% as.character()` [@ribeiro2016; @pedersen2019]

@@

Large phenotypic differences were found between the sites by supervised random forest (Supplementary  figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`). 
This was as a results of differences in descriptors relating tree size and likely the ages of the trees between the sites. 
Post adjustment the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.

@@

A significant difference was found in the PDI between non-symptomatic (control) and symptomatic (AOD and COD) decline statuses using an independant two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
Significance was also found in the DAI between the AOD and COD groups (`r glue("t({DAI_groups_ttest$parameter}) = {signif(DAI_groups_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(DAI_groups_ttest$estimate1,digits = 3)` and `r signif(DAI_groups_ttest$estimate2,digits = 3)` for the AOD and COD groups respectively.
A scatter plot of PDI against DAI is shown in Figure # for all 174 trees.
It can clearly be seen that the

@@

- Accessability of collecting descriptors
- Suitability of collected phenotypic descriptors for describing decline status
- Concession of model interpretability to predictive capability by using a machine learning approach
- Trends of key phenotypic descriptors withing the PDI and DAI measures
- Sensitivity provided for integration with large scale omics analyses as well as long term monitoring for forest condition
- Index thresholds for category definitions (healthy, declining, severe decline etc.)
- Semi-supervised can be used if unsupervised model is unable to pull out decline severity scale
- Scalability of method for wider application
- When viewed in the 2D continuum, the categorisation of AOD to only include bleeds of certain bacteria seems likely to be too specific and by definition should only describe the period of time under which the tree have been declining.
Perhaps subcategories of AOD would be more suitable such as Agrilus associated AOD or Armillaria associated AOD. 
- Thresholds for decline groupings within index ranges
- Validation of decline scores - chicken and paradox
- Refined definition of remission
- use of unsupervised over supervised analysis

@@

- Crown condition descriptors most important for describing overall decline severity in the PDI.
- Bacterial bleed related descriptors are most important for describing decline acuteness.
- Decline indexes provide intuitive measures of decline severity based on simple phenotypic measurements that do not require specialist equipment

>>>>

---
title: "Index measures for oak decline severity using phenotypic descriptors"
output: 
  #bookdown::word_document2:
  # toc: false
  bookdown::pdf_document2: 
    toc: false
    number_sections: false
bibliography: references.bib
---

@@

**Keywords:** acute oak decline, chronic oak decline, phenotyping, machine learning, unsupervised learning, random forest

@@

- Site types
- Field phenotyping protocol
- Crown assessments
- Other assessments
- Descriptions of collected phenotypic descriptors

@@

Descriptors were collected relating to the trees shape, size, crown condition and the presence of biotic agents including insects, bacteria and fungi.
These descriptors are shown in Table #.
Crown radius, canopy closure and tap tests for dead stem tissue were measured for each cardinal point.
The crown radius was
The AOD stem symptoms (active bleeds, black staining, callused wounds) were collected by measuring longest axis of each individual bleed, stain or callused wound found within the basal 3m of the stem (ref).
The frequency of *A. biguttatus* exit holes were counted along with the presence or absence of each of oval shaped exit holes, small circular shaped exit holes and ground level fungal fruiting bodies were recorded from the basal 2m of the stem (ref). 
All trees were manually assigned a decline status category; control, AOD, COD or remission.

All raw data and analysis code used to generate this manuscript are available at https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors.

@@

- Descriptors converted to presence and absence
- Aggregation of cardinal observations.
- Symptom type sizes average and frequency taken
- DBH converted to m and symptom sizes to mm

R v`r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and subsequent analysis of phenotypic data [@R2019].

@@

A total of 31 raw descriptors collected

### Composite descriptor calculation

Seven additional, composite descriptors were calculated to account for the three dimensional attributes of the tree and covariate nature of certain descriptors as suggested by @zarnoch2004.
This then gave a total of 38 raw and composite phenotypic descriptors.
The additional descriptors are defined as follows:

@@

***Agrilus* exit hole density (m^-2^)** - density of observed *Agrilus* exit holes across the surveyed 2m basal trunk area.

$$e = \frac{h}{\frac{2d}{100}\pi}$$

$h$ = number of *Agrilus* exit holes, $d$ = diameter at breast height (cm)

@@

To account for these differences and to make the trees across the sites comparable in the context of decline, an adjustment of site means to the overall mean was applied to each of the size descriptors (diameter at breast height, lower crown height, timber height, total height, crown radius). 
An example of this adjustment is shown for diameter at breast height in Supplementary Figure # and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table #.

@@

The phenotypic descriptors were analysed using 100 repetitions of unsupervised random forest using the randomForest R package v`r packageVersion('randomForest') %>% as.character()` [@liaw2002].
The proximity matrices from each repetition were averaged to a single proximity matrix ($P$).
This was then transformed to a dissimilarity matrix ($1 - P$) and scaled to two dimensions using  multidimensional scaling (MDS).
The first MDS dimension was min-max scaled to give a scale between 0 and 1 with more severely declining trees having a score closer to 1.
This dimension was named the PDI.
The second MDS dimension ($D$) was first inverted ($1 - D$) to ensure the correct dimension orientation relative to decline type.
The dimension was then min-max scaled and then transformed ($2D - 1$) to give a scale between -1 and 1 with COD trees having a score closer to -1 and AOD having a score closer to 1. 
This dimension was named the DAI.
With different data, it is likely that the MDS dimensions may require inversion prior to min-max scaling to ensure their correct orientation.

The code used to generate the decline indexes is available as an R package pdi v`r packageVersion('pdi') %>% as.character()` at https://github.com/jasenfinch/pdi.

@@

Individual random forest regression models of the phenotypic descriptors were trained, supervised using each decline index.
Optimal number of trees (ntree) and number randomly of variables (mtry) parameters for random forest were tuned prior to training. 
The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and as follows: PDI - `r glue('ntree = {PDI_rf_tune_params$ntree}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)}%')`; DAI - `r glue('ntree = {DAI_rf_tune_params$ntree}, mtry = {DAI_rf_tune_params$mtry}, PMAE = {DAI_rf_tune_params$PMAE %>% signif(digits = 3)}%')` (Supplementary figure #)

To understand the descriptors contributing to each of the indexes, Local Interpretable Model-Agnostic Explanations (LIME) using the lime R package v`r packageVersion('lime') %>% as.character()` [@ribeiro2016; @pedersen2019]

@@

Large phenotypic differences were found between the sites by supervised random forest (Supplementary  figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`). 
This was as a results of differences in descriptors relating tree size and likely the ages of the trees between the sites. 
This adjustment removed inter-site variability but maintained intra-site variability, allowing comparisons to be made between the trees at the different sites in the context of decline severity.
Post adjustment the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.

@@

Trees with a score around 0 are neither distinctively AOD or COD sharing characteristics of both conditions in trees with PDI greater than 0.3.
A significant difference was found in the PDI between non-symptomatic (control) and symptomatic (AOD and COD) decline statuses using an independant two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
Significance was also found in the DAI between the AOD and COD groups (`r glue("t({DAI_groups_ttest$parameter}) = {signif(DAI_groups_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(DAI_groups_ttest$estimate1,digits = 3)` and `r signif(DAI_groups_ttest$estimate2,digits = 3)` for the AOD and COD groups respectively.
A scatter plot of PDI against DAI is shown in Figure # for all 174 trees.
It can clearly be seen that the

@@

- Accessability of collecting descriptors
- Suitability of collected phenotypic descriptors for describing decline status
- Concession of model interpretability to predictive capability by using a machine learning approach
- Trends of key phenotypic descriptors withing the PDI and DAI measures
- Sensitivity provided for integration with large scale omics analyses as well as long term monitoring for forest condition
- Index thresholds for category definitions (healthy, declining, severe decline etc.)
- Semi-supervised can be used if unsupervised model is unable to pull out decline severity scale
- Scalability of method for wider application
- Thresholds for decline groupings within index ranges
- Validation of decline scores - chicken and paradox
- Refined definition of remission
- use of unsupervised over supervised analysis
- Scale inversion likely need with new data

@@

- Crown condition descriptors most important for describing overall decline severity in the PDI.
- Bacterial bleed related descriptors are most important for describing decline acuteness.
- Decline indexes provide intuitive measures of decline severity based on simple phenotypic measurements that do not require specialist equipment
- Encouraged to use this approach in other contexts
